---
title: "Guidelines to create PPI plots from polar volume files in bioRad"
author: 'Liesbeth Verlinden, Peter Desmet, Stijn Van Hoey'
date: "January 2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating PPI plots}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r include=FALSE}
library(bioRad, quietly = TRUE)
```

[BioRad](https://github.com/adokter/bioRad#biorad) provides code to plot a PPI of an [ODIM](http://eumetnet.eu/activities/observations-programme/current-activities/opera/) scan. An ODIM scan can be extracted from an ODIM polar volume (pvol), though at moment of writing this documentation, single scans from folders cannot yet be read in. 

Here is example code how to extract a scan from a polar volume:
```{r}
# locate example polar volume
pvol.path <- system.file("extdata", "volume.h5", package="bioRad")
# load the pvol:
pvol <- read.pvol(pvol.path)
# copy the first scan to a new object 'scan2'
scan2 <- pvol$scans[[1]]
scan2
scan2$params[[4]]
```

This is example code how to create a PPI plot:
```{r}
# load an example scan:
data(SCAN)
# make ppi for the scan
ppi <- ppi(SCAN)

# plot the first scan parameter of the ppi, which in this case is "VRADH"
plot(ppi)
# plot the reflectivity quantity
plot(ppi,param="DBZH")
# change the range of reflectivities to plot to -30 to 50 dBZ
plot(ppi,param="DBZH",zlim=c(-30,50))
```

Additionally different types of underlying maps can be added using the function `basemap`:
```{r message=FALSE}
# grab a basemap that matches the extent of the ppi
basemap <- basemap(ppi)
# map the reflectivity quantity of the ppi onto the basemap
map(ppi,map=basemap,param="DBZH")

# download a different type of basemap, e.g. satellite imagery
      ### see get_map() in ggmap library for full documentation of options ###
basemap <- basemap(ppi,maptype="satellite")
# map the radial velocities onto the satellite imagery
map(ppi,map=basemap,param="VRADH")
```

More functionallity of the package bioRad, such as plotting vertical profiles of birds (VPB), adjusting the radar cross section and calculating migration traffic rates, can be find in its documentation.
