---
title: "Guidelines to load bird profiles into bioRad"
author: 'Liesbeth Verlinden, Peter Desmet, Stijn Van Hoey'
date: "January 2017"
output: html_document
---

```{r include=FALSE}
library(bioRad, quietly = TRUE)
library(ggmap, quietly = TRUE)
```

This document gives an introduction how to access and load weather radar bird profiles into the R package [bioRad](https://github.com/adokter/bioRad#biorad). This process is done in two steps:

1. Downloading the data.
2. Loading the data into bioRad.

## 1. Downloading bird profile data

The bird profiles, generated by the [vol2bird algorithm](https://github.com/adokter/vol2bird#vol2bird-algorithm), are available in this [data repository](http://enram.github.io/infrastructure/). The repository is structured in the following way:
```{}
    └──
      country
        └── radar
              └── year
                    └── month
                          └── day
                               └── hour
                                     └── filename: dkrom_vp_20170114231500.h5
```

The name of the file consist of three parts:

* Country and radar code: the first two letters give the country, the last three the radar. The abbrevation is based on the [ODIM code](http://eumetnet.eu/wp-content/themes/aeron-child/observations-programme/current-activities/opera/database/OPERA_Database/index.html). 
* vp: abbrevation of vertical profile.
* Time and date: YYYYMMDDHHMMSS.

### Manually downloading bird profile data

There are two ways to manually download the profiles:

* One by one.
* Each radar has per month a zip file with all the data of that month. This zip file is placed on the same level as the month folder (so in the year folder).

When unpacking the zipfiles after download, the data will be structured in the same hierarchy of folders as on the repository: 
```{}
  └── month
        └── day
             └── hour
                   └── filename: dkrom_vp_20170114231500.h5
```

### BioRad bird profile download function

The `download_vp` function provided with the BioRad function supports the download of a set of zip files, defined by a start date, end date, country and radar. As en example, downloading the available data of Octobre and November 2016 for two Belgian radars to the current working directory:

```{r downloadvp}
download_vp("2016-10-01", "2016-11-30", c("be"), 
            c("jab", "wid"), localpath = ".")
```

## 2. Loading bird profile data into bioRad
Once the data is downloaded in your folder, you have two ways to load the data into bioRad. 

* As a single file.
* As several single files at the same time.

Below is some example code:
```{r eval=FALSE}
# read single bird profile
vp <- read.vp("PATH/TO/FILE.h5")
# read several single bird profiles
vp.list <- read.vplist("PATH/TO/FILE1.h5", "PATH/TO/FILE2.h5", "PATH/TO/FILE3.h5")
```

An additional function `retrieve_vp_paths` will allow you to browse through the data folders and select many paths to bird profiles based on country, radar, and a specified time frame, which then can be passed to the `read.vplist` function. 

The abbrevations of country and radar are the same as used in the ODIM code, see explanation under Section 1. Downloading bird profile data.

Below are examples of which combinations of arguments you can use to select specific bird profiles.

### General
If no arguments are speficied, the paths of _all vp files in de folders_ will be added to the list:
```{r eval=FALSE}
# read paths of all vp files in the folders
list.paths <- retrieve_vp_paths(path="PATH/TO/DATA-DIR")
```

### Location
Only specifying the country selects the paths of the vp files of _all radars of that country_:
```{r eval=F}
# read paths of all vp files of all radars in Belgium
list.paths <- retrieve_vp_paths(path="PATH/TO/DATA-DIR", country="be")
```

Specifying the radar selects _all vp files of that radar_:
```{r eval=FALSE}
# read paths of all vp files of the radar 'Jabbeke'
list.paths <- retrieve_vp_paths(path="PATH/TO/DATA-DIR", radar="jab")
```

You can also specify _several radars within a country_:
```{r eval=FALSE}
# read paths of all vp files of the French radars Momuy, Nancy and Bordeau
list.paths <- retrieve_vp_paths(path="PATH/TO/DATA-DIR", radar=c("mom","nan","bor"))
```

You CANNOT specify _all radars via the country_ argument and then add radars _of another country_ by the radar argument:
```{r eval=FALSE}
# read paths of all vp files af all Belgian radars and the French radars Momuy, Nancy and Bordeau
list.paths <- retrieve_vp_paths(path="PATH/TO/DATA-DIR", country="be", radar=c("mom","nan","bor"))
## WRONG!! ##
```


### Time
You can specify a time frame using the start_date and end_date arguments, which takes integer values in the following format: "YYYY-MM-DD". See the example below:
```{r eval=FALSE}
# read paths of all vp files of all Swiss radars from Nov/2016 till Jan/2017
list.paths <- retrieve_vp_paths(path="PATH/TO/DATA-DIR", country="cz", start_date="2016-11-01", end_date="2017-01-31")
```

*Special case*: if you want the paths to profiles for a _certain month across all years_, you will have to call the `retrieve_data_range` fuction several times across the years and put the outcomes together:
```{r eval=FALSE}
# read paths of all vp files of all Swiss radars for Nov/16
list.paths16 <- retrieve_vp_paths(path="PATH/TO/DATA-DIR", country="cz", start_date="2016-11-01", end_date="2016-11-31")
# read paths of all vp files of all Swiss radars for Nov/17
list.paths17 <- retrieve_vp_paths(path="PATH/TO/DATA-DIR", country="cz", start_date="2017-11-01", end_date="2017-11-31")
# put paths together 
list.path <- c(list.paths16, list.paths17)
```

Once the list with paths is created, it can be added to the `read.vplist` function to create a list of VP's:
```{r eval=F}
#make list of bird profiles with list of their paths
vp.list2 <- read.vplist(list.paths)
```


## 3. Introduction to plotting in bioRad
The bioRad package allows to plot PPI's, VP's and VPB's.

### VP
BioRad also allows to plot the bird profiles, see example code below:
```{r}
# load in an example bird profile
data(VP)
# plot bird profile
plot(VP,line.col='blue')
```

### VPB plots
The package provides a function to plot vertical profiles of birds (VPB) too. This function takes in time series of bird profiles, which can be generated by the function `vpts`:
```{r eval=FALSE}
# make time series from all vertical profiles in a list
vpts.be <- vpts(vp.list2)
# make time series from all vertical profiles of the 'Jabbeke' radar
vpts.jab <- vpts(vp.list2, radar="jab")
```

Note that, to be able to plot the time series, these need to be regular. The function `regularize` regularizes the times series on a grid, with option to specify the time interval:
```{r eval=FALSE}
# regularize the time series on a 5 minute interval grid
vpts.jab.reg <- regularize(vpts.jab, interval=15)
```

Below is example code of how to make a VPB plot:
```{r}
# locate example file
VPtable <- system.file("extdata", "VPtable.txt", package="bioRad")
# load and regularize time series of vertical profiles
ts <- regularize(readvp.table(VPtable,radar="KBGM", wavelength='S'))

# plot density of individuals for the first 500 time steps, in the altitude layer 0-3000 m.
plot(ts[1:500], ylim=c(0,3000))
# plot total reflectivity factor (rain,birds,insects together):
plot(ts[1:500], ylim=c(0,3000), quantity="DBZH")
```

More functionallity of the package bioRad, such as adjusting the radar cross section and calculating migration traffic rates, can be find in its documentation.